---
title: "Exploratory Data Analysis"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Gene Expression

Let us start by loading the gene expression in long format, i.e. one row per
gene × sample combination.

```{r message=FALSE}
library(tidyverse)
library(OSD758)

(counts <- gene_expression())
```

### Most expressed genes

Calculating median counts per gene:

```{r}
median_gene_expression <-
  counts |>
  dplyr::group_by(ensembl_gen_id, gene_symbol) |>
  dplyr::summarise(median_counts = median(counts), .groups = "drop") |>
  dplyr::arrange(desc(median_counts))

median_gene_expression
```

The mostly expressed gene is [rhodopsin
(Rho)](https://www.informatics.jax.org/marker/MGI:97914) with a median of over
2.7 million reads, perhaps not so surprising given that these RNAseq samples are
from retina.

### Not expressed genes

On the contrary we have also genes that are not expressed at all. Actually,
one-third of all genes (19,028) show no expression whatsoever in mouse retina:

```{r}
total_gene_expression <-
  counts |>
  dplyr::group_by(ensembl_gen_id, gene_symbol) |>
  dplyr::summarise(total_counts = sum(counts), .groups = "drop") |>
  dplyr::arrange(desc(total_counts))

total_gene_expression |>
  dplyr::filter(total_counts == 0L)
```

### Gene expression distribution

The gene expression distribution is extremely skewed as one-third have no
expression at all.

```{r}
median_gene_expression |>
  ggplot(aes(x = median_counts)) +
  geom_density()
```

Removing genes with a median expression of zero counts still produces a
very skewed distribution.

```{r}
median_gene_expression |>
  dplyr::filter(median_counts != 0) |>
  ggplot(aes(x = median_counts)) +
  geom_density()
```

Log-scale transforming the median counts allows to conciliate very low counts
with million count values, we can appreciate (roughly) three intervals:

- Below 100 counts
- Between 100 and 10,000 counts
- Above 10,000 counts

```{r}
median_gene_expression |>
  dplyr::filter(median_counts != 0) |>
  ggplot(aes(x = log10(median_counts))) +
  geom_density()
```
